---
title: "Chapter 4. EDA with R"
output: html_notebook
---

I am walking through the [Chapter 4: RDA with R of *Statistics for Soil Survey*](http://ncss-tech.github.io/stats_for_soil_survey/chapters/4_exploratory_analysis/4_exploratory_analysis.html) by Tom D’Avello and Stephen Roecker to freshen up my R skills.

So following Tukey. The domain here is some soil sample stuff or something, I don't know anything about it. Exciting.

Let's first install the interface package, whose [CRAN description](https://cran.r-project.org/package=soilDB).

> A collection of functions for reading data from USDA-NCSS soil databases.

```{r install package}
install.packages('soilDB')
```

## 3. Data Inspection

```{r}
library(soilDB)
```

The book uses loafercreek dataset from the CA630 Soil Survey Area.

```{r loaddata}
data("loafercreek")
```

```{r}
# generalized horizon designations
n <- c("A", "BAt", "Bt1", "Bt2", "Cr", "R")
# REGEX ruled
p <- c("A", "BA|AB", "Bt|Bw", "Bt3|Bt4|2B|C", "Cr", "R")
```

Compute genhz labels - whatever that means - and add to `loafercreek` dataset. `generalize.hz` is from `aqp`, a dependency of the `soilDB` package.

```{r}
loafercreek$genhz <- generalize.hz(loafercreek$hzname, n, p)
```

Extract the horizon table, part of the `loafercreek` dataset, using a special function for.

```{r extract horizon table}
h <- horizons(loafercreek)
```

Examine the matching pairing of the genhz label to the hzname

```{r}
table(h$genhz, h$hzname)
```

And also

```{r}
View(h)
```

Statistical overview of all the columns, though the book only summarizes some of the columns to save space.

```{r}
summary(h)
```

Ok so `summary()` gives different kinds of output depending on data type of each column in the `h` dataframe, as well as missing values if they are present.

Check levels of a variable. Why does this return `NULL`? levels ought to work for factors.

```{r}
levels(h$genhz)
```
This is suitable for factor or character variables.

```{r}
sort(unique(h$hzname))
```

To replace values, e.g. typos, we could do

    h$hzname <- ifelse(h$hzname == "BT", "Bt", h$hzname)

or

    h$hzname[h$hzname == "BT"] <- "Bt"

Which for this example is a question of taste, but maybe more complicated say computations might be useful to do the first style.



## 4 Descriptive Statistics

We will look at 

| Statistic | Function        |
| ----------| ----------------|
| Mean      | `mean()`        |
| Median    | `median()`      |
| Mode      | `sort(table(), decreasing = TRUE)[1]`
| Standard deviation | `sd()` |
| Quantiles | `quantile()`    |

### 4.1 Measures of Central Tendency

We could call these *representative values* in "NASIS speak", the *Statistics for Soil Survey* says.

Mean is formally $\bar{x} = \frac{\sum^n_{i=1} x_i}{n}$ for $n$ samples, specifically using samples here and not population, hence $\bar{x}$. Dropping missing values

```{r clay mean}
mean(na.exclude(h$clay)) # or mean(h$clay, na.rm = TRUE)
```

Median is the middle, ie 50th percentile.
```{r clay median}
median(h$clay, na.rm = TRUE)
```

And finally mode, the most frequent measurement. R provides no function for mode, but we can

```{r clay mode}
sort(table(round(h$clay)), decreasing = TRUE)[1]
```

Ok so why two values? Because those two are equally frequent.

For frequencies we can use `table()` and `xtabs()`.

```{r frequency of genhz values}
table(h$genhz)
```

To see two qualitative (nominal or ordinal) variables

```{r frequency of genhz and texcl values}
table(h$genhz, h$texcl)
```

Also with margins ie. sums

```{r frequency of genhz values w. margins}
addmargins(table(h$genhz, h$texcl))
```

Proportions relative to the rows (`margin = 1`, below) or columns (`margin = 2`)

```{r proportions}
round(prop.table(table(h$genhz, h$texture_class), margin = 1) * 100)
```

Aggregations, here mean clay by genhz

```{r}
aggregate(clay ~ genhz, data = h, mean)
```

### 4.2 Measures of Dispersion

Variance, ie deviation from mean is $s^2 = \frac{\sum_{i=1}^n(x_i - \bar{x})^2}{n-1}$ aka *sum of squares*.

```{r clay variance}
var(h$clay, na.rm = TRUE)
```

Standard deviation is $s = \sqrt{s^2}$, which is a silly way to say it.

```{r clay tandard deviation}
sd(h$clay, na.rm = TRUE)
```

```{r clay quantiles}
quantile(h$clay, na.rm = TRUE)
```

Range ie. min and maximum

```{r clay range}
range(h$clay, na.rm = TRUE)
```

and value of range.

```{r clay range value}
diff(range(h$clay, na.rm = TRUE))
```

Interquantile range (IQR) is the range from the upper 75% quartile to the lower 25% quartile, ie. the 50% of observations occurring in the mid-range of the sample.

```{r clay iqr}
IQR(h$clay, na.rm = TRUE)
```

### 4.3 Correlation

```{r middle horizon depth}
h$hzdepm <- (h$hzdepb + h$hzdept) / 2
```

```{r clay selected correlations}
vars <- c("hzdepm", "clay", "sand", "total_frags_pct", "phfield")
round(cor(h[vars], use = "complete.obs"), 2)
```

