---
title: "Chapter 4. EDA with R"
output: html_notebook
---

I am walking through the [Chapter 4: RDA with R of *Statistics for Soil Survey*](http://ncss-tech.github.io/stats_for_soil_survey/chapters/4_exploratory_analysis/4_exploratory_analysis.html) by Tom D’Avello and Stephen Roecker to freshen up my R skills.

So following Tukey. The domain here is some soil sample stuff or something, I don't know anything about it. Exciting.

Let's first install the interface package, whose [CRAN description](https://cran.r-project.org/package=soilDB).

> A collection of functions for reading data from USDA-NCSS soil databases.

```{r install package}
install.packages('soilDB')
```

## 3. Data Inspection

```{r}
library(soilDB)
```

The book uses loafercreek dataset from the CA630 Soil Survey Area.

```{r loaddata}
data("loafercreek")
```

```{r}
# generalized horizon designations
n <- c("A", "BAt", "Bt1", "Bt2", "Cr", "R")
# REGEX ruled
p <- c("A", "BA|AB", "Bt|Bw", "Bt3|Bt4|2B|C", "Cr", "R")
```

Compute genhz labels - whatever that means - and add to `loafercreek` dataset. `generalize.hz` is from `aqp`, a dependency of the `soilDB` package.

```{r}
loafercreek$genhz <- generalize.hz(loafercreek$hzname, n, p)
```

Extract the horizon table, part of the `loafercreek` dataset, using a special function for.

```{r extract horizon table}
h <- horizons(loafercreek)
```

Examine the matching pairing of the genhz label to the hzname

```{r}
table(h$genhz, h$hzname)
```

And also

```{r}
View(h)
```

Statistical overview of all the columns, though the book only summarizes some of the columns to save space.

```{r}
summary(h)
```

Ok so `summary()` gives different kinds of output depending on data type of each column in the `h` dataframe, as well as missing values if they are present.

Check levels of a variable. Why does this return `NULL`? levels ought to work for factors.

```{r}
levels(h$genhz)
```
This is suitable for factor or character variables.

```{r}
sort(unique(h$hzname))
```

To replace values, e.g. typos, we could do

    h$hzname <- ifelse(h$hzname == "BT", "Bt", h$hzname)

or

    h$hzname[h$hzname == "BT"] <- "Bt"

Which for this example is a question of taste, but maybe more complicated say computations might be useful to do the first style.



## 4 Descriptive Statistics

We will look at 

| Statistic | Function        |
| ----------| ----------------|
| Mean      | `mean()`        |
| Median    | `median()`      |
| Mode      | `sort(table(), decreasing = TRUE)[1]`
| Standard deviation | `sd()` |
| Quantiles | `quantile()`    |

### 4.1 Measures of Central Tendency

We could call these *representative values* in "NASIS speak", the *Statistics for Soil Survey* says.

Mean is formally $\bar{x} = \frac{\sum^n_{i=1} x_i}{n}$ for $n$ samples, specifically using samples here and not population, hence $\bar{x}$. Dropping missing values

```{r clay mean}
mean(na.exclude(h$clay)) # or mean(h$clay, na.rm = TRUE)
```

Median is the middle, ie 50th percentile.
```{r clay median}
median(h$clay, na.rm = TRUE)
```

And finally mode, the most frequent measurement. R provides no function for mode, but we can

```{r clay mode}
sort(table(round(h$clay)), decreasing = TRUE)[1]
```

Ok so why two values? Because those two are equally frequent.

For frequencies we can use `table()` and `xtabs()`.

```{r frequency of genhz values}
table(h$genhz)
```

To see two qualitative (nominal or ordinal) variables

```{r frequency of genhz and texcl values}
table(h$genhz, h$texcl)
```

Also with margins ie. sums

```{r frequency of genhz values w. margins}
addmargins(table(h$genhz, h$texcl))
```

Proportions relative to the rows (`margin = 1`, below) or columns (`margin = 2`)

```{r proportions}
round(prop.table(table(h$genhz, h$texture_class), margin = 1) * 100)
```

Aggregations, here mean clay by genhz

```{r}
aggregate(clay ~ genhz, data = h, mean)
```

### 4.2 Measures of Dispersion

Variance, ie deviation from mean is $s^2 = \frac{\sum_{i=1}^n(x_i - \bar{x})^2}{n-1}$ aka *sum of squares*.

```{r clay variance}
var(h$clay, na.rm = TRUE)
```

Standard deviation is $s = \sqrt{s^2}$, which is a silly way to say it.

```{r clay tandard deviation}
sd(h$clay, na.rm = TRUE)
```

```{r clay quantiles}
quantile(h$clay, na.rm = TRUE)
```

Range ie. min and maximum

```{r clay range}
range(h$clay, na.rm = TRUE)
```

and value of range.

```{r clay range value}
diff(range(h$clay, na.rm = TRUE))
```

Interquantile range (IQR) is the range from the upper 75% quartile to the lower 25% quartile, ie. the 50% of observations occurring in the mid-range of the sample.

```{r clay iqr}
IQR(h$clay, na.rm = TRUE)
```

### 4.3 Correlation

```{r middle horizon depth}
h$hzdepm <- (h$hzdepb + h$hzdept) / 2
```

```{r clay selected correlations}
vars <- c("hzdepm", "clay", "sand", "total_frags_pct", "phfield")
round(cor(h[vars], use = "complete.obs"), 2)
```

## 5. Graphical Methods

Here, we are going to look at bars, histograms, density, Q-Q plots, box-whiskers and scatter & line methods. Quantile-Quantile (Q-Q) plot is "a plot of the actual data against a normal distribution".

There is some talking about "ggplot geoms", but that is maybe different from `ggplot2`. The ggplot geoms are used here, see below. No, `ggplot()` is a function from `ggplot2`, described as

> `ggplot()` initializes a ggplot object

```{r}
require("ggplot2")
```


### 5.1 Distributions

Bar plot, histogram and density plot.

```{r}
idx <- is.na(h$texture_class)
ggplot(h[!idx, ], aes(x = texcl)) +
  geom_bar()
```

Note how the `h` is indexed with `[!indx, ]`, is by selecting rows, and the comma in there select all columns. For `ggplot()` to map aesthetics, there's the `aes()` mapping.

Histogram

```{r clay histogram}
ggplot(h, aes(x = clay)) +
  geom_histogram(bins = nclass.Sturges(h$clay))
```

and a density curve too, aka kernel density plot. Compared to a histogram which gives a count or percentage of observations, the preferred density estimation gives instead probability of observing any given value. AUC = 1.

```{r clay density curve}
ggplot(h, aes(x = clay)) +
  geom_density()
```

Boxplot gives distribution with quantiles ie. 25%, 50% and 75% percentiles, mix and max values and outliers. The *Statistics for Soil Survey* says the whiskers are typically 1.5 * IQR above or below the 3rd and 1st quartile, and thus defines what an outlier is. Ok stay vigilant to how whisker plots are defined.

I wasn't familiar with the term IQR, but makes sense and is a useful concept.

```{r clay boxplot per genhz with whiskers}
ggplot(h, aes(x = genhz, y = clay)) +
  geom_boxplot()
```

Ok so quantile-quantile (Q-Q) plot is a plot of the actual data values against a normal distribution, with mean of 0 and standard deviation of 1. So data against theoretical.

```{r clay qqplot}
ggplot(h, aes(sample = clay)) +
  geom_qq()
```

```{r}
ggplot(h, aes(sample = total_frags_pct)) +
  geom_qq()
```

> If the data set is perfectly symmetric (i.e. normal), the data points will form a straight line. Overall this plot shows that our clay example is more or less symmetric. However the second plot shows that our rock fragments are far from evenly distributed.
>
> A more detailed explanation of QQ plots may be found on Wikipedia:
https://en.wikipedia.org/wiki/QQ_plot

some discussion about properties of normalcy and that it is often used and assumed. Median is more robust than mean.

```{r clay scatterplot with hzperm}
ggplot(h, aes(x = clay, y = hzdepm)) +
  geom_point() +
  ylim(100, 0)
```

```{r clay lines with hzperm by peeid}
ggplot(h, aes(y = clay, x = hzdepm, group = peiid)) +
  geom_line() +
  coord_flip() +
  xlim(100, 0)
```

SPLOM with `GGally`

```{r pairs}
install.packages("GGally")
library(GGally)

vars <- c("hzdepm", "clay", "phfield", "total_frags_pct")
ggpairs(h[vars])
```

3rd dimension with color, shape, si, layers et cetera.

```{r}
idx <- h$genhz %in% c("Cr", "R", "not-used")
```

```{r clay vs hzperm by genhz}
ggplot(h[!idx, ], aes(x = clay, y = hzdepm, color = genhz)) +
  geom_point() +
  ylim(100, 0)
```

```{r}
ggplot(h[!idx, ], aes(x = clay, color = genhz)) +
  geom_density()
```

```{r}
ggplot(h, aes(x = genhz, fill = texture_class)) +
  geom_bar()
```

```{r sites heatmap}
ggplot(site(loafercreek), aes(x = landform_string, y = pmkind)) + 
  geom_tile(alpha = 0.2) 
```

Then some facets, first with boxplot

```{r facet boxplot}
df <- reshape2::melt(h,
                     id.vars = c("peiid", "genhz", "hzdepm"),
                     measure.vars = c("clay", "phfield", "total_frags_pct"))

idx <- df$genhz %in% c("Cr", "R", "not-used")

ggplot(df[!idx, ], aes(x = genhz, y = value)) +
  geom_boxplot() +
  xlab("genhz") +
  facet_wrap(~ variable, scales = "free_y")
```

```{r}
library(aqp)

s <- slice(loafercreek, 1:100 ~ clay + phfield + total_frags_pct)
s <- slab(s, fm = ~ clay + phfield + total_frags_pct, 
          slab.fun = function(x) quantile(x, c(0.1, 0.5, 0.9), na.rm = TRUE)
          )

ggplot(s, aes(x = top, y = X50.)) +
  geom_line() +
  geom_ribbon(aes(ymin = X10., ymax = X90., x = top), alpha = 0.2) +
  xlim(c(100, 0)) +
  coord_flip() +
  facet_wrap(~ variable, scales = "free_x")
```

